---
title: "Gov 50 Final Project"
author: "Ashley Adirika"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Milestone 1: Initial Project Thoughts

I am interested in exploring data related to the incarceration of certain demographic groups-- namely looking at trends that exist for marginalized and economically disadvantaged groups. 



## Milestone 2: Data Source and Question Proposal 


```{r}
load("/Users/ashleyadirika/Downloads/HawesSPPQ_ReplicationData.RData")

FinalProjectData <- table

```



The question that my final project will seek to answer is “What is the relationship between Black/White poverty inequality and the Black/White incarceration ratio in the United States?” I hypothesize that in states with high poverty inequality between white and Black populations, there is a disproportionate number of Black people incarcerated compared to their white counterparts. 

To test this hypothesis, I will be using data published by Daniel Hawes entitled “Replication Data for: Social Capital, Racial Context and Incarcerations in the American States."  In doing so, I plan to take the average Black/White poverty rate for each state provided in the dataset across the thirty-year data collection period. The average Black/White poverty rate will serve as my independent variable. Then, I will look at the average Black/White incarceration ratio for these states. To contextualize my understanding of these values, I will also utilize demographic information in the dataset that concerns state population size (and the racial breakdowns of these populaces) and total prison populations (and their racial breakdowns).  


## Milestone 3

```{r}
 
library(tidyverse)
library(dplyr)
library(ggplot2)

FinalProjectData

##Independent Variable is average Black/ White poverty rate (bwpovratio)

## Dependent Variable is average Black/White Incarceration Ratio (prison_BWratio)


Data_Visual_1 <- ggplot(data = FinalProjectData, 
                       mapping = aes(x = bwpovratio, 
                                     y = prison_BWratio)) + 
  geom_point() + 
  geom_smooth( method = "lm", se = FALSE) +
  scale_x_log10() + 
  labs( x = "Average Poverty Inequality (Black/White Ratio)",
        y = "Average Black/White Incarceration Ratio",
        title = "Poverty Inequality and Incarceration Ratio",
        subtitle = "Across the United States", 
        caption = "Source: Hawes, Daniel, 2017")

  
Data_Visual_1





```


