---
title: "Gov 50 Final Project"
author: "Ashley Adirika"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

##Introduction: 

The question that my final project will seek to answer is “What is the relationship between the amount of Black state legislatures in office and the Black/White Prison Ratio across states in 2008?” I hypothesize that the more Black state legislatures are in office for a given state, the lower the Black/White Prison Ratio will be. 

This is my hypothesis based on my assumption that states with more Black elected officials will have less tangible prejudice against Black populations – given that they are being elected into office. That assumption then leads me to believe that Black populations would not be over-policed in comparison to white populations within the state. 

This is a very important area of research given that in the status quo, Black people are over-policed compared to their white counterparts. This project may present a starting point for potential solvency. That is, if this query determines that there is a significant and strong negative correlation between Black representation in the state legislature and Black/White prison ratios then it would be important to uncover why. With an understanding that correlation does not mean causation, additional research and tests could be conducted to determine whether electing more Black representation to state legislature might meaningful decrease the Black prison population in America. 

To test my hypothesis, I used data published by Daniel Hawes entitled “Replication Data for: Social Capital, Racial Context and Incarcerations in the American States."

This data was pulled from a longitudinal study, that examined the “differential effects of social capital on policy equity in state outcomes” (Hawes, 2017) by looking at several variables like unemployment, black/white poverty ratios, black/white incarceration ratios in each state from 1980 until 2010. For my final project, I decided to calculate the average percentage of Black State Legislatures in Office and the average Black/White Prison Ratio from each state from 1980 until 2010 and look at the relationship between those values. This became my base data with which I conducted all of my analyses. It is important to note that the District of Columbia had missing data for both of these variables from 1980 until 2010, so DC will not be represented in my analysis. 

```{r}
library(readr)
library(gapminder)
library(tidyverse)
library(dplyr)
library(ggplot2)

```


```{r}
load("/Users/ashleyadirika/Downloads/HawesSPPQ_ReplicationData.RData")

FinalProjectData <- table
```



Project Data: Data from all Years - Grouped by State 
```{r}

state_summary <- FinalProjectData |> 
  group_by(statename) |> 
  summarize( 
    avg_blk_leg = mean(blk_leg, na.rm = TRUE),
    avg_prison_BWratio = mean(prison_BWratio, na.rm = TRUE)
  ) |> 
  drop_na(avg_blk_leg, avg_prison_BWratio)

state_summary 


  


```


##Data section: 

In order to get a base look at what the relationship between the average percentage of Black State legislatures in Office and the average Black/White Prison Ratio from rach state from 1980 - 2010 was, I created a smoothed curve to summarize the relationship between the two variables. 


Summarize/Visualize the Dependent Variable - Smoothed Curve
```{r}

ggplot(data = state_summary,
       mapping = aes(x = avg_blk_leg,
                     y = avg_prison_BWratio)) +
  geom_point() + geom_smooth()



```

Based on this preliminary view of the data, there appears to be an overall negative relationship between the two variables. However, the trend appears to increase a bit before decreasing. This is perhaps an early indication that the relationships between the two variables are not incredibly strong. 



##Results


```{r}

Data_Visual_1.6 <- ggplot(data = state_summary, 
                       mapping = aes(x = avg_blk_leg, 
                                     y = avg_prison_BWratio)) + 
  geom_point() + 
  geom_smooth( method = "lm", se = FALSE) +
  scale_x_log10() + 
  labs( x = "Average Percent African Americans in State Legislatures",
        y = "Average Black/White Incarceration Ratio",
        title = "Relationship Between Black State Legislatures and Black/White Incarceration by State",
        subtitle = "Across 50 US States from 1980-2010", 
        caption = "Source: Hawes, Daniel, 2017") +
   scale_x_log10()

  
Data_Visual_1.6 


```

Calculating Correlation (Between Avg State Legislature + Avg Black/White Incarceration)

```{r}

cor(state_summary$avg_blk_leg, state_summary$avg_prison_BWratio,
    use = "pairwise")




```

The correlation between the average percentage of African Americans in State Legislatures and the Average Black/White Incarceration Ratio is approximately -0.206. This means that there is a very weak negative correlation between the two variables. In other words, as the amount of African Americans in State Legislatures increases, the Average Black/White Incarceration Ratio tends to decrease, however there is significant variability in the data such that this trend is not strong. 


After getting this value, I conducted multiple regressions to hold the average percentage of the black population of each state constant as well as the diversity index. 




```{r}

statesummary_blkpop <- FinalProjectData |> 
  group_by(statename) |> 
  summarize( 
    avg_blk_leg = mean(blk_leg, na.rm = TRUE),
    avg_prison_BWratio = mean(prison_BWratio, na.rm = TRUE), 
    avg_pop_pctblk = mean(pop_pctblk, na.rm = TRUE), 
    avg_racialdiversity = mean(racialdiversity, na.rm = TRUE)
  ) |> 
  drop_na(avg_blk_leg, avg_prison_BWratio, avg_pop_pctblk, avg_racialdiversity)

statesummary_blkpop


```




```{r}



pop.fit <- lm(avg_prison_BWratio ~ avg_blk_leg +  avg_pop_pctblk + avg_racialdiversity, 
              data = statesummary_blkpop)

summary(pop.fit)


```


The intercept ( 9.8304) represents the estimated average black/white prison ratio when the independent variables are held at 0. 

The avg_blk_leg coefficient of (0.3161) represents the estimated change in the average black/white prison ratio for a percentage average increase in the black state legislature. 

The avg_pop_pctblk coefficient of (-0.2303) represents the estimated change in the average black/white prison ratio for a percentage average increase in the black population of a state. 

The avg_racialdiversity coefficient of (-5.3025) represents the estimated change in the average black/white prison ratio for an average unit increase in the racial diversity index of a state. 

None of the coefficients are statistically significant given that their p-vales are not less than 0.05.



Since the relationship between avg_blk_leg and black/white incarceration ratio flipped in the multiple regression, it led me to suspect that the flip was due to the avg_blk_leg  and avg_pop_pctblk variables being highly correlated. To test this theory, I tested the correlation between avg_blk_leg and avg_pop_pctblk. 


```{r}
cor(statesummary_blkpop$avg_blk_leg, statesummary_blkpop$avg_pop_pctblk,
    use = "pairwise")
```



After running a correlation test between the Average Percentage of Black State Legislatures and the Average Black population percentage, my suspicion that the values were highly correlated was confirmed. The correlation between the two variables is strong and positive (0.9395601). As such, I wanted to evaluate whether there was a synergistic effect between "avg_pop_pctblk" and "avg_racialdiversity". To do so, I ran another multiple regression --- this time, however, I multiplied "avg_pop_pctblk" and "avg_racialdiversity". 


```{r}

pop.fit_2 <- lm(avg_prison_BWratio ~ avg_blk_leg :  avg_pop_pctblk + avg_racialdiversity, 
              data = statesummary_blkpop)

summary(pop.fit_2)



```


Based on the multiple regression, the coefficient for the interaction term is not statistically significant. This is because the p-value (0.229) is greater than 0.05. Thus, despite the variables (avg_blk_leg and avg_pop_pctblk) being highly correlated, the data does not suggest a synergistic effect. However, based on this multiple regression the coefficients can be evaluated as such: 

The intercept coefficient (9.623729) represents the estimated average black/white prison ratio when the independent variables are held at 0.  
The avg_racialdiversity coefficient (-4.032670) represents the estimated average black/white prison ratio increase when the racial diversity index increases by one unit. 

The avg_blk_leg:avg_pop_pctblk coefficient (-0.004002) represents the estimated average black/white prison ratio increase when both avg_blk_leg and avg_pop_pctblk are present/ together increase by a unit.   

None of the values are statistically significant. This means that we cannot conclude that the null hypothesis is rejected -- that is, we fail to reject the null hypothesis --- that the selected variables (Average Black State Legislatures & Average Black Population Percentage, Average Racial Diversity Index) have no effect on average Black/White prison ratios. 





##Conclusion: 

Ultimately, the data demonstrates that there is a weak negative correlation between Black representation in the state legislature and Black/White prison ratios. When I ran the multiple regression table to control for the Black population and Diversity Index, the coefficients were not statistically significant. This is because the p-values were all less than 0.05. 

With this understanding, we cannot interpret the data as causal. That is, the presence of more Black State Legislatures does not cause a decrease in Black/White incarceration ratios. Instead, the data indicates that there is a merely weak correlation. In addition to what has already been considered, a potential confounding variable is the ideology of voters in a given state. That is, if voters are more “tough on crime” in one state, it is likely to impact the amount of people incarcerated on the whole. If the voters in this same state also hold racist or anti-diversity sentiments, it is likely that they have larger Black/White incarceration ratios. This is because voters elect officials who create policy that is supported by their constituents. This is something that the current data I am working with, cannot account for and thus, a limitation. 

Given more time and money to conduct this research, I would have liked to find data about the dominant voter ideologies of each state I was working with. I could leverage this information to control for the prior-mentioned confounding variables, or even observe beyond speculation what the relationship between voter ideology and Black/white incarceration actually is. 



